A Reduced Order Model for Porous Media Flows: Coupling Linear Regression with Principal Component Analysis (PCA) using SVD or High-Order Singular Value Decomposition (HOSVD) in Upscaling Calculations for Permeability in Reservoir Simulations

### 2018-12-03 Papers/AppliedMaths2018/Resubmission/Stochastic_Upscaling_UK_AMM_2018_TrackedRevision.tex
Created Tracked version of updated manual

%%graphis package for simple commands
%% \usepackage{graphics}
%\usepackage{cases}
%% or use the graphicx package for more complicated commands
\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
\usepackage{epsfig}
%\usepackage{subfig}
\usepackage{comment}

@@

%\title{A Reduced Order Model for Porous Media Flows which Couples Linear Regression with Principal Component Analysis (PCA) using SVD or HOSVD: Upscaling the Permeability Field for Multiphase Reservoir Flow Simulation}
\title{A Reduced Order Model for Porous Media Flows: Coupling Linear Regression with Principal Component Analysis (PCA) using SVD or High-Order Singular Value Decomposition (HOSVD) in Upscaling Calculations for Permeability in Reservoir Simulations}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes}
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering, University of Aberdeen, UK}

@@

Representation of heterogeneous petro-physical properties (\eg permeability) within a large system, is a challenge that cuts across many industries that deal with flows in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domains. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all available data values. This challenge gave birth to upscaling, a model order reduction method.

In this paper, a single realization of a permeability dataset was upscaled by first projecting the permeability field into a principal component analysis space (using SVD and HOSVD for 2- and 3-D geometries, respectively). Then linear regression was employed within each of the principal component spaces to obtain approximate representations of the initial principal component. The new principal components are then recombined to form a new permeability dataset with a reduced order model. This PCA-linear regression method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits/performance.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Naturally occuring rock formations are inherently heterogeneous, which means that geophysical properties, such as permeability, pore spaces (\ie porosity) and pore throat vary spatially at different length scales. Heterogeneity of rock formations is relevant in calculations of transport and storage of fluids in porous media (\ie multiphase flow in subsurface rocks). Therefore, the representation of such heterogeneous properties is of great importance to oil and gas (\ie reservoir management), carbon capture and storage (CCS, \ie CO$_2$ transportation and storage in subsurface rocks) and waste management (\ie remediation of contaminated soil).

In essence, rock heterogeneity makes it impossible to obtain reliable and accurate spatial information about geological properties throughout the domain (\ie uncertainty problem). Additionally, due to limitations in computing resources, it is neither efficient nor possible to use exact values of geological properties at all spatial coordinates of the domain to perform fine-grid simulations (\ie scaling problem) \cite{chen_2006,miller_1998,Renard_1997}. This is particularly true for large geological domains which span several kilometres. It is possible to overcome these challenges by a family of techniques called {\it upscaling}. Upscaling techniques solve both uncertainty and scaling problems by replacing discrete geological and fluid properties of detailed high-resolution domains with coarse descriptions (\ie low-resolution) of these properties \cite{Vereecken_2007}.

Upscaling techniques which best preserve statistical geophysical and thermo-physical properties of high-resolution domains result in optimal prediction of fluid flow dynamics. Over the past 60 years, several upscaling techniques were developed (see \cite{Hasting_2001, Renard_1997, Szymkiewicz_2013}). Traditional upscaling techniques are deterministic in nature, however over the last decade, techniques classed as stochastic have received great attention from academic and industrial porous media communities worldwide \cite{Guilleminot_2012, Ravalec-Dupin_2010, Verwoerd_2009}. One of the reasons for this is because stochastic upscaling techniques are robust enough to handle multiphase flow in porous media and they are able to address scaling and uncertainity problems.statistical

\medskip
{\it Upscaling} is a term commonly used in the oil and gas sector and is often referred, throughout applied mathematics and engineering communities, as a form of model order reduction (or reduced order models, MOR/ROM). Depending on the context in which MOR is used, it can be simply referred as `dimensionality reduction'. The definition of MOR also depends on the context in which it is used, but it suffices to state that MOR's are ale to captures essential features of a system, structure or parameter \cite{Schilders2008}. This implies that an original system exists with more detailed information. Figure~\ref{fig:IllustrationMOR} presents an interpretation of MOR concept in which a graphical representation of a rabbit is reconstructed with a reduced number of facets (rhs of this illustration).

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998, which introduced the passive reduced-order interconnect macromodeling algorithm (PRIMA). Truncated balanced realization is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis (PCA) \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and proper orthogonal decompositions (POD) \cite{Sirovich_1987} are also MOR methods discussed in the references published in 1984 and 1987, respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA.

\bigskip
 In this work, upscaling is performed through linear regression in the PCA space which is obtained with high-order singular value decomposition (HOSVD/SVD) in 2-/3-D problems. This work focuses on reducing the order of permeability distribution for porous media flow simulations.

\bigskip
A brief overview of current upscaling methods in reservoir simulation is described in Section~\ref{section:general_upscaling}. Section~\ref{section:mathematical_model} introduces the mathematical model used to describe the relevant physics for multi-fluid flow through porous media and highlights the relevance of the permeability field. The mathematical model is discretised by a high-order accurate control volume finite element method (CVFEM) \cite{Gomes_2017}. Pre-processing statistical properties and post-simulation multiphase flow behaviour are investigated to assess the various upscaling techniques used. Four upscaling techniques are investigated in this work and they are named arithmetric mean, harmonic mean, probability density function (PDF) and PCA (using SVD and HOSVD methods) coupled with linear regression. In Section~\ref{section:pca_linreg}, fundamental mathematics of SVD and associated properties are presented. Then, the concept of spaces is explained to give the reader the understanding necessary to appreciate the transformations and projections involved in this work. Finally, Section~\ref{section:pca_linreg} describes the reduced order model reduction for the permeability field within the principal component space. Section~\ref{section:model_simulation} is dedicated to model description and simulation. It starts by describing the high-resolution problem (\ie `base case') in which the four upscaling techniques are applied to. A brief summary of the pre-processing steps for each of the models is also presented. Section~\ref{section:results_discussion} provides the results and further discussion of the results, it also gives further interpretation to the result obtained from the SVD upscaling technique. Finally, conclusions are drawn in Section~\ref{section:conclusion}.

@@

Upscaling methods replace discrete geological and fluid properties of detailed (high-resolution) domains with coarse descriptions (low-resolution) of these properties \cite{Vereecken_2007}. Upscaling of heterogeneous properties has attracted the interest of several porous media flow communities (\eg oil and gas, CCS and waste management) worldwide for over $7$ decades. Those interested in large-scale reservoirs or groundwater generally take a two-stage geo-statistical approach.

@@

\subsection{PCA, SVD, POD, KLT and Reservoir Characterisation}\label{section:reservoir_char}

PCA, SVD and POD are closely related. In the petroleum engineering community where these techniques are used for reservoir characterization, they are sometimes used interchangable, and are normally referred to as Karhunen-Loeve Transformations (KLT) \cite{Jafarpour_2009}.

A number of recent research in reservoir engineering focus on  history matching (HM) problems (see \cite{Afra_2013, Insuasty_2017, Jafarpour_2009, Tavakoli_2010, Xiao_2018}). History matching is the process of building or adjusting one or more sets of reservoir configurations (or realisations), such that reservoir models are able to closely reproduces past flow behaviour (\eg flow rate, production volume, pressure etc) in a geological formation, aiming to predict future flow behaviour with good accuracy \cite{Rwechungura_2011}. It is important to note that reservoir configurations could be either a numerical model representing the system of parameters (\ie{variables}) which characterise the reservoir, or simply a model for parameterizing one of more components (\ie{variables}) of the system.

\citet{Jafarpour_2009} compares KLT and discrete cosine transform (DCT) and concluded that the DCT is computationally more efficient than the KLT while being almost as accurate. \citet{Tavakoli_2010} noted that HM calculation of sensitivities for all production data with respect to system parameters is not feasible. Hence, they explored a new parameterization for reducing the number of variables based on the principal right singular vectors of the dimensionless sensitivity matrix.

Perhaps, work performed by \citet{Afra_2013,Afra_2016,Afra_2014} and \citet{Insuasty_2017} are more closely related to the one performed here. However, similar to the work of \citet{Jafarpour_2009}, they are more related to comparing several realizations than reducing the dimensionality of the dataset for a single realization. The work of \citet{Xiao_2018} is also closely related to that performed here and interestingly, they used the same flow simulator model used in this work. However, similarly to \citet{Tavakoli_2010}, \citeauthor{Xiao_2018} focused on reducing the number of variable used to describe the system.

Essentially, recent developments in the field of reservoir simulation described in this section make use of SVD theory to select the most suitable realization from multiple realizations or to reduce the representative system variables. However, the work described here is concerned with reducing dimensionality of the dataset for a single realization of a single system variable (\ie{permeability}).

@@

\medskip
The discretised saturation and Darcy equations are solved using a multigrid-like approach as described by \citeauthor{pavlidis2016}~\cite{pavlidis2016}. The numerical formulation is fully described by \citeauthor{Gomes_2017}~\cite{Gomes_2017} (see also \cite{adam_2016, salinas2015}). These numerical methods are embedded in the next-generation flow simulator Fluidity/IC-FERST model software\footnote{\href{http://multifluids.github.io}{http://multifluids.github.io}} (a full description of the model can be found in \cite{Gomes_2017, jackson_2013,salinas_2018}). Although, the numerical formulation used here is relatively new, it is not the focus of this work. The result and discussion section (Section \ref{section:results_discussion}) refers to \citet{dawe_2008} for validation of the numerical formulation with respect to the 2D simulation. However, extensive model and software quality assurance (\ie verification and validation tests) can be found in the aforementioned references \cite{adam_2016, Gomes_2017,jackson_2013, pavlidis2016, salinas2015}.

@@

\section{Coupled PCA and Linear Regression}\label{section:pca_linreg}

@@

\subsubsection{Understanding the permeability space}\label{subsubsection:visualization_permspace}
Permeability can be represented as either tensor or scalar quantity, however for the purpose of the following explanation (and the remaining of this work), permeability is assumed as a scalar quantity. In addition, in the numerical simulations performed for this work, permeability is discretised element-wise with projection to the control volume space within each element-pairs (see \cite{Christou_2018}). Permeability values are assigned to the centre of the FE $\left(\text{\ie P}_{0}\text{DG element-pair}\right)$. For simplicity, in the remaining part of this sub-section, a $2D$ geometry is assumed which is discretized with triangular elements (\PN[1]{2} and \PNDG[1]{1} element-pairs). This explanation allows one to easily visualize this permeability space because it is defined by Cartesian coordinate system.

@@

\subsubsection{Understanding the principal component spaces}\label{subsubsection:visualization_pcspaces}
Essentially, the existing permeability space is transformed (with SVD and HOSVD for 2- and 3-D permeability data, respectively) to principal component spaces (PCS). Within the PCS, traditional PCA dimensionality reduction can be optionally performed (\ie{feature selection or elimination}, Section~\ref{subsubsection:pca_dimred_linreg}), before interpolation is used to implement further data reduction in PCS. The reduced data in the PCS is then transformed back into the permeability space to obtain a reduced permeability space. An algorithm which impliments the procedure described here is presented in Section~\ref{subsubsection:svdcase_preprocess_algorithm}

The PCS is a particularly challenging to visualize as it represents projections which are embedded in SVD/HOSVD factors of the original permeability dataset. An aid for visualizing the PCS for a 2D permeability data is to write out $\mathbf{U}$, $\mathbf{\Sigma}$ and $\mathbf{V^{\intercal}}$ matrices (or SVD factors) and consider the singular vectors (\ie columns in $\mathbf{U}$ and rows in $\mathbf{V^{\intercal}}$) as separate PCS. A similar procedure can be extended to HOSVD for 3D permeability dataset. The decomposition of a matrix into its factors using SVD \cite{Cline_2007, Tharwat_2016} or HOSVD \cite{Kolda_2009, Lathauwer_2000} is covered in details in several published materials. However, for completeness, \ref{subsection:svd_brief} and \ref{subsection:hosvd_brief} contain brief overviews of SVD and HOSVD, respectively.

\subsubsection{Reduced Order Model/Principal Component Space Reduction}\label{subsubsection:pca_dimred_linreg}
For the PCA-Linear regression reduced order model (PCA-LinReg) of the permeability data, all work is performed in PCS. This can be conducted in two steps (for 2-D models), thus:
\begin{description}
  \item[Step 1:] Perform PCA \cite{Hotelling_1933, Tharwat_2016} using the SVD method, where the singular values are examined and, based on a pre-defined criteria, $z$ singular values are selected, starting from $\sigma_{1}$ and in a decreasing order, where $z \leq r$. Note that, in the case with $z = r$, it can be assumed that {\it Step 1} has been ignored. However, when $z < r$, the first $z$ columns of $\mathbf{U}$ and the first $z$ rows of $\mathbf{V^{\intercal}}$ are extracted from $\mathbf{U}$ and $\mathbf{V^{\intercal}}$ to form new matrices $\mathbf{U_{new}}$ and $\mathbf{V^{\intercal}_{new}}$ respectively. The selected $z$ singular values also form a new diagonal matrix, $\mathbf{\Sigma_{new}}$ .

  \item[Step 2:] Interpolation within each singular vector (column in $\mathbf{U}$ and row in $\mathbf{V^{\intercal}}$) is conducted to reduce the number of data points in each column of $\mathbf{U}$ and in $\mathbf{V^{\intercal}}$. From this step,  $\mathbf{U_{new2}}$ and $\mathbf{V^{\intercal}_{new2}}$ are obtained. The reduced permeability field is calculated by multiplying  $\mathbf{U_{new2}}$, $\mathbf{\Sigma_{new}}$ and $\mathbf{V^{\intercal}_{new2}}$.

@@

These two steps can be readily replicated for 3-D models using HOSVD to obtain the PCA space.

Finally on the PCS, it is worth noting that two methods exist for obtaining PCA space: co-variance and SVD methods (see \cite{Tharwat_2016} detailed description of both methods). In this work, the authors have opted to use the SVD method to obtain the PCA space, as singular vectors which make up the PCS have strong correlation to the coordinate axis they represent. This was exploited in making the assumption that a linear relationship exist for the data in each of the PC spaces, hence the linear regression performed to reduce the data in the PCS.

@@

\citet{AndrewNg_2018} was quoted to have said ``PCA is not linear regression, and despite some cosmetic similarity, these are actually totally different algorithms''. The main differences between these two mathematical methods is summarised in the Table~\ref{table:pca_vs_linreg} and can also be deduced from Figure~\ref{fig:PCA_is_not_LinReg}.

\begin{table}[h!]
\centering
\begin{tabular}{|m{1em} |m{14em} |m{14em}|} 
 \hline
  & \begin{center}{\bf PCA}\end{center} & \begin{center}{\bf Linear Regression}\end{center} \\ [0.5ex] 
 \hline
 1 & Determines the best vector (or set of vectors) which compresses the data points by minimizing the orthogonal distance to the points & Approximates the best fit for the points by minimizing the error \\
 \hline
 2 & Essentially, it forms a projection of the original points onto the new vector space & Predicts $y$ from $x$ through a function $f(x_1,x_2,....,x_n) = y$ \\
 \hline
 3 & Not used for predictions, as original feature space (\ie $x_1$ and $x_2$) give way to a new reduced dimension. Cordinates $x_1, x_2,...,x_n$ give way to $z_1,....,z_k$ where $k \leq n$ & \\[1ex]
 \hline
\end{tabular}
\caption{Table comparing PCA with Linear Regression.}
\label{table:pca_vs_linreg}
\end{table}

@@

The author's future work will provide a quantitative analysis of the results obtained from the coupled PCA-Linear regression method. Additionally, future work will present other methods which follow a similar principle but also seek to optimize the quantitative results obtained.

>>>>

%%graphis package for simple commands
%% \usepackage{graphics}
%\usepackage{cases}
%% or use the graphicx package for more complicated commands
\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
\usepackage{epsfig}
%\usepackage{subfig}
\usepackage{comment}
\usepackage[normalem]{ulem}

@@

%\title{A Reduced Order Model for Porous Media Flows which Couples Linear Regression with Principal Component Analysis (PCA) using SVD or HOSVD: Upscaling the Permeability Field for Multiphase Reservoir Flow Simulation}
\title{\red{\sout{A Reduced Order Model for Permeability Fields Using Singular Value Decomposition}}\blue{A Reduced Order Model for Porous Media Flows: Coupling Linear Regression with Principal Component Analysis (PCA) using SVD or High-Order Singular Value Decomposition (HOSVD) in Upscaling Calculations for Permeability in Reservoir Simulations}}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes}
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering, University of Aberdeen, UK}

@@

Representation of heterogeneous properties (\ie permeability in this case) within a large system, is a challenge that cuts across many industries that deal with flows in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domains. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all available data values. This challenge gave birth to upscaling, a model order reduction method.

In this paper, a single realization of a permeability field was upscaled by first projecting the permeability field into a principal component analysis space (using SVD for 2D field and HOSVD for 3D fields). Then linear regression was employed within each of the principal component spaces to obtain approximate representations of the initial principal component. The new principal components are then recombined to for a new permeability field with a reduced order. This PCA-linear regression method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits/performance.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Naturally occuring rock formations are inherently heterogeneous, which means that \red{\sout{rock}} \blue{geophysical} properties, such as permeability, pore spaces (\ie porosity) and pore throat vary spatially at different length scales. Heterogeneity of rock formations is relevant in calculations of transport and storage of fluids in porous media (\ie multiphase flow in subsurface rocks). Therefore, the representation of such heterogeneous properties is of great importance to oil and gas (\ie reservoir management), carbon capture and storage (CCS, \ie CO$_2$ transportation and storage in subsurface rocks) and waste management (\ie remediation of contaminated soil).

In essence, rock heterogeneity makes it impossible to obtain reliable and accurate spatial information about geological properties throughout the domain (\ie uncertainty problem). Additionally, due to limitations in computing resources, it is neither efficient nor possible to use exact values of geological properties at all spatial coordinates of the domain to perform fine-grid simulations (\ie scaling problem) \cite{chen_2006,miller_1998,Renard_1997}. This is particularly true for large geological domains which span several kilometres. It is possible to overcome these challenges by a family of techniques called {\it upscaling}. Upscaling techniques solve both uncertainty and scaling problems by replacing discrete geological and fluid properties of detailed high\blue{-}resolution domains with coarse descriptions (\ie low\blue{-}resolution) of these properties \cite{Vereecken_2007}.

Upscaling techniques which best preserve statistical \red{\sout{and flow dynamics behaviour of the high resolution domain give the best representation of heterogeneous properties. There are several upscaling techniques which fall under different categories}} \blue{geophysical and thermo-physical properties of high-resolution domains result in optimal prediction of fluid flow dynamics. Over the past 60 years, several upscaling techniques were developed (see} \cite{Hasting_2001, Renard_1997, Szymkiewicz_2013}). Traditional upscaling techniques are deterministic in nature, however over the last decade, techniques classed as stochastic have received great attention from academic and industrial porous media communities worldwide \cite{Guilleminot_2012, Ravalec-Dupin_2010, Verwoerd_2009}. One of the reasons for this is because stochastic upscaling techniques are robust enough to handle multiphase flow in porous media and they are able to address scaling and uncertainity problems.statistical

\medskip
{\it Upscaling} is a term commonly used in the oil and gas sector and is often referred, throughout \red{\sout{the}} applied mathematics and engineering communities, as a form of model order reduction (or reduced order models, MOR/ROM). Depending on the context in which MOR is used, it can \blue{be} simply \red{\sout{be}} referred as `dimensionality reduction'. The definition of MOR also depends on the context \red{\sout{it is being}} \blue{in which it is} used, but it suffices to state that \red{\sout{MOR captures}} \blue{MOR's are able to capture} the essential features of a system, structure or parameter \cite{Schilders2008}. This implies that an original system exists with more detailed information. Figure~\ref{fig:IllustrationMOR} presents an interpretation of MOR concept in which a graphical representation of a rabbit is reconstructed with a reduced number of facets (rhs of this illustration).

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998, which introduced the passive reduced-order interconnect macromodeling algorithm (PRIMA). Truncated balanced realization is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis \blue{(PCA)} \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and proper orthogonal decompositions \blue{(POD)} \cite{Sirovich_1987} are also MOR methods discussed in the references published in 1984 and 1987, respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA. \red{\sout{In this work, upscaling is performed by SVD in principal}}

\bigskip
 \blue{In this work, upscaling is performed through linear regression in the PCA space which is obtained with high-order singular value decomposition (HOSVD/SVD) in 2-/3-D problems.} This work focuses on reducing the order of permeability distribution for porous media flow simulations.

\red{\sout{Four upscaling techniques are investigated in this work, arithmetic and harmonic means are used to obtain the first two upscaled representation and these two are classed as deterministic techniques. The third is a randomly generated permeability field with Gaussian distribution, prescribed by a probability density function (PDF) which was obtained from a domain discretised with high-resolution mesh with known permeability distribution (\ie base case). The fourth upscaling technique is the crux of this research, it introduces singular value decomposition (SVD) as a method of upscaling using the concept of principal component space to reduce the order of the permeability distribution.}}

\bigskip
\blue{A brief overview of current upscaling methods in reservoir simulation is described in Section~\ref{section:general_upscaling}.} Section~\ref{section:mathematical_model} introduces the mathematical model used to describe the relevant physics for multi-fluid flow through porous media and highlights the relevance of the permeability field. The mathematical model is discretised by a high-order accurate control volume finite element method (CVFEM) \cite{Gomes_2017}. Pre-processing statistical properties and post-simulation multiphase flow behaviour are investigated to assess the various upscaling techniques used. Four upscaling techniques are investigated in this work and they are named arithmetric mean, harmonic mean, probability density function (PDF) and PCA (using SVD and HOSVD methods) coupled with linear regression. In Section~\ref{section:pca_linreg}, fundamental mathematics of SVD and associated properties are presented. Then, the concept of spaces is explained to give the reader the understanding necessary to appreciate the transformations and projections involved in this work. Finally, Section~\ref{section:pca_linreg} describes the reduced order model reduction for the permeability field within the principal component space. Section~\ref{section:model_simulation} is dedicated to model description and simulation. It starts by describing the high-resolution problem (\ie `base case') in which the four upscaling techniques are applied to. A brief summary of the pre-processing steps for each of the models is also presented. Section~\ref{section:results_discussion} provides the results and further discussion of the results, it also gives further interpretation to the result obtained from the SVD upscaling technique. Finally, conclusions are drawn in Section~\ref{section:conclusion}.

@@

\red{\sout{Section~\ref{section:intro} briefly introduced the rational behind upscaling}} Upscaling methods replace discrete geological and fluid properties of detailed (high-resolution) domains with coarse descriptions (low-resolution) of these properties \cite{Vereecken_2007}. Upscaling of heterogeneous properties has attracted the interest of several porous media flow communities (\eg oil and gas, CCS and waste management) worldwide for over $7$ decades. Those interested in large-scale reservoirs or groundwater generally take a two-stage geo-statistical approach.

@@

\subsection{\blue{PCA, SVD, POD, KLT and Reservoir Characterisation}}\label{section:reservoir_char}

\blue{PCA, SVD and POD are closely related. In the petroleum engineering community where these techniques are used for reservoir characterization, they are sometimes used interchangable, and are normally referred to as Karhunen-Loeve Transformations (KLT)} \cite{Jafarpour_2009}.

\blue{A number of recent research in reservoir engineering focus on  history matching (HM) problems (see} \cite{Afra_2013, Insuasty_2017, Jafarpour_2009, Tavakoli_2010, Xiao_2018}). \blue{History matching is the process of building or adjusting one or more sets of reservoir configurations (or realisations), such that reservoir models are able to closely reproduces past flow behaviour (\eg flow rate, production volume, pressure etc) in a geological formation, aiming to predict future flow behaviour with good accuracy} \cite{Rwechungura_2011}. \blue{It is important to note that reservoir configurations could be either a numerical model representing the system of parameters (\ie{variables}) which characterise the reservoir, or simply a model for parameterizing one of more components (\ie{variables}) of the system.}

\citet{Jafarpour_2009}\blue{ compares KLT and discrete cosine transform (DCT) and concluded that the DCT is computationally more efficient than the KLT while being almost as accurate.} \citet{Tavakoli_2010} \blue{noted that HM calculation of sensitivities for all production data with respect to system parameters is not feasible. Hence, they explored a new parameterization for reducing the number of variables based on the principal right singular vectors of the dimensionless sensitivity matrix.}

\blue{Perhaps, work performed by} \citet{Afra_2013,Afra_2016,Afra_2014} and \citet{Insuasty_2017}\blue{ are more closely related to the one performed here. However, similar to the work of} \citet{Jafarpour_2009}, \blue{they are more related to comparing several realizations than reducing the dimensionality of the dataset for a single realization. The work o}f \citet{Xiao_2018} \blue{is also closely related to that performed here and interestingly, they used the same flow simulator model used in this work. However, similarly to} \citet{Tavakoli_2010}, \citeauthor{Xiao_2018} \blue{focused on reducing the number of variable used to describe the system.}

\blue{Essentially, recent developments in the field of reservoir simulation described in this section make use of SVD theory to select the most suitable realization from multiple realizations or to reduce the representative system variables. However, the work described here is concerned with reducing dimensionality of the dataset for a single realization of a single system variable (\ie{permeability}).}

@@

\medskip
The discretised saturation and Darcy equations are solved using a multigrid-like approach as described by \citeauthor{pavlidis2016}~\cite{pavlidis2016}. The numerical formulation is fully described by \citeauthor{Gomes_2017}~\cite{Gomes_2017} (see also \cite{adam_2016, salinas2015}). These numerical methods are embedded in the next-generation flow simulator Fluidity/IC-FERST model software\footnote{\href{http://multifluids.github.io}{http://multifluids.github.io}} (a full description of the model can be found in \cite{Gomes_2017, jackson_2013,salinas_2018}).\blue{ Although, the numerical formulation used here is relatively new, it is not the focus of this work. The result and discussion section (Section} \ref{section:results_discussion}) refers to \citet{dawe_2008} \blue{for validation of the numerical formulation with respect to the 2D simulation. However, extensive model and software quality assurance (\ie verification and validation tests) can be found in the aforementioned references} \cite{adam_2016, Gomes_2017,jackson_2013, pavlidis2016, salinas2015}.

@@

\section{\red{Singular Value Decomposition (SVD) and Application for Reducing the Order of Permeability Fields}\blue{Coupled PCA and Linear Regression}}\label{section:pca_linreg}

\subsection{\red{A Brief Introduction to SVD}}
\red{Given a matrix $\mathbf{A}\in\mathcal{R}^{m\times n}$, the singular value decomposition (SVD) is a method for factorizing $\mathbf{A}$, such that,
\begin{equation}
  \mathbf{A} = \mathbf{U}\mathbf{\Sigma}\mathbf{V}^{\intercal}.
\end{equation}}

@@

\subsubsection{Understanding the permeability space}\label{subsubsection:visualization_permspace}
\blue{Permeability can be represented as either tensor or scalar quantity, however for the purpose of the following explanation (and the remaining of this work), permeability is assumed as a scalar quantity. In addition, in the numerical simulations performed for this work, permeability is discretised element-wise with projection to the control volume space within each element-pairs (see \cite{Christou_2018}).} Permeability values are assigned to the centre of the FE $\left(\text{\ie P}_{0}\text{DG element-pair}\right)$. For simplicity, in the remaining part of this sub-section, a $2D$ geometry is assumed which is discretized with triangular elements (\PN[1]{2} and \PNDG[1]{1} element-pairs). This explanation allows one to easily visualize this permeability space because it is defined by Cartesian coordinate system.

@@

\subsubsection{Understanding the principal component spaces}\label{subsubsection:visualization_pcspaces}
\blue{Essentially, the existing permeability space is transformed (with SVD and HOSVD for 2- and 3-D permeability data, respectively) to principal component spaces (PCS). Within the PCS, traditional PCA dimensionality reduction can be optionally performed (\ie{feature selection or elimination}, Section~\ref{subsubsection:pca_dimred_linreg}), before interpolation is used to implement further data reduction in PCS. The reduced data in the PCS is then transformed back into the permeability space to obtain a reduced permeability space. An algorithm which impliments the procedure described here is presented in Section~\ref{subsubsection:svdcase_preprocess_algorithm}}

\blue{The PCS is a particularly challenging to visualize as it represents projections which are embedded in SVD/HOSVD factors of the original permeability dataset. An aid for visualizing the PCS for a 2D permeability data is to write out $\mathbf{U}$, $\mathbf{\Sigma}$ and $\mathbf{V^{\intercal}}$ matrices (or SVD factors) and consider the singular vectors (\ie columns in $\mathbf{U}$ and rows in $\mathbf{V^{\intercal}}$) as separate PCS.\blue{ A similar procedure can be extended to HOSVD for 3D permeability dataset. The decomposition of a matrix into its factors using SVD} \cite{Cline_2007, Tharwat_2016} or HOSVD }\cite{Kolda_2009, Lathauwer_2000} is covered in details in several published materials. However, for completeness, \ref{subsection:svd_brief} and \ref{subsection:hosvd_brief} contain brief overviews of SVD and HOSVD, respectively.

\subsubsection{Reduced Order Model/Principal Component Space Reduction}\label{subsubsection:pca_dimred_linreg}
\blue{For the PCA-Linear regression reduced order model (PCA-LinReg) of the permeability data, all work is performed in PCS. This can be conducted in two steps (for 2-D models), thus}:
\begin{description}
  \item[Step 1:] Perform PCA \cite{Hotelling_1933, Tharwat_2016} \blue{using the SVD method, where the singular values are examined and, based on a pre-defined criteria, $z$ singular values are selected, starting from $\sigma_{1}$ and in a decreasing order, where $z \leq r$. Note that, in the case with $z = r$, it can be assumed that {\it Step 1} has been ignored. However, when $z < r$, the first $z$ columns of $\mathbf{U}$ and the first $z$ rows of $\mathbf{V^{\intercal}}$ are extracted from $\mathbf{U}$ and $\mathbf{V^{\intercal}}$ to form new matrices $\mathbf{U_{new}}$ and $\mathbf{V^{\intercal}_{new}}$ respectively. The selected $z$ singular values also form a new diagonal matrix, $\mathbf{\Sigma_{new}}$ .}

  \item[Step 2:] \blue{Interpolation within each singular vector (column in $\mathbf{U}$ and row in $\mathbf{V^{\intercal}}$) is conducted to reduce the number of data points in each column of $\mathbf{U}$ and in $\mathbf{V^{\intercal}}$. From this step,  $\mathbf{U_{new2}}$ and $\mathbf{V^{\intercal}_{new2}}$ are obtained. The reduced permeability field is calculated by multiplying  $\mathbf{U_{new2}}$, $\mathbf{\Sigma_{new}}$ and $\mathbf{V^{\intercal}_{new2}}$.}

@@

\blue{These two steps can be readily replicated for 3-D models using HOSVD to obtain the PCA space.}

\blue{Finally on the PCS, it is worth noting that two methods exist for obtaining PCA space: co-variance and SVD methods (see} \cite{Tharwat_2016} \blue{detailed description of both methods). In this work, the authors have opted to use the SVD method to obtain the PCA space, as singular vectors which make up the PCS have strong correlation to the coordinate axis they represent. This was exploited in making the assumption that a linear relationship exist for the data in each of the PC spaces, hence the linear regression performed to reduce the data in the PCS. }

@@

\citet{AndrewNg_2018} \blue{was quoted to have said ``PCA is not linear regression, and despite some cosmetic similarity, these are actually totally different algorithms''. The main differences between these two mathematical methods is summarised in the Table~\ref{table:pca_vs_linreg} and can also be deduced from Figure~\ref{fig:PCA_is_not_LinReg}.}
\blue{
\begin{table}[h!]
\centering
\begin{tabular}{|m{1em} |m{14em} |m{14em}|} 
 \hline
  & \begin{center}{\bf PCA}\end{center} & \begin{center}{\bf Linear Regression}\end{center} \\ [0.5ex] 
 \hline
 1 & Determines the best vector (or set of vectors) which compresses the data points by minimizing the orthogonal distance to the points & Approximates the best fit for the points by minimizing the error \\
 \hline
 2 & Essentially, it forms a projection of the original points onto the new vector space & Predicts $y$ from $x$ through a function $f(x_1,x_2,....,x_n) = y$ \\
 \hline
 3 & Not used for predictions, as original feature space (\ie $x_1$ and $x_2$) give way to a new reduced dimension. Cordinates $x_1, x_2,...,x_n$ give way to $z_1,....,z_k$ where $k \leq n$ & \\[1ex]
 \hline
\end{tabular}
\caption{Table comparing PCA with Linear Regression.}
\label{table:pca_vs_linreg}
\end{table}
}

@@

\blue{The author's future work will provide a quantitative analysis of the results obtained from the coupled PCA-Linear regression method. Additionally, future work will present other methods which follow a similar principle but also seek to optimize the quantitative results obtained.}
